---
title: "Getting started with rcds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rcds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rcds)
```

`rcds` is the **Richard Cartographic Design System** — an opinionated toolkit
that turns a recurring map style into a reusable, consistent, accessible system.
This vignette is the working tour: the parts that need no spatial data run live;
the map-building parts are shown as code you copy into a real script.

The full written standard lives in `docs/RCDS.md`; this is the hands-on path.

## 1. The shape of a map

Every published RCDS map follows the same five-step skeleton:

```r
library(rcds)
library(ggplot2)

rcds_fonts("editorial")                 # 1. pick ONE font voice

main <- ggplot(my_sf) +                 # 2. build the map
  geom_sf(aes(fill = value), colour = NA) +
  scale_fill_rcds_c("seq_blue") +       # 3. an accessible palette
  labs(
    title    = "US School District Enrollment",
    subtitle = "ACS 5-Year Estimates",
    caption  = rcds_signature(          # 4. the standardized signature
      challenge = "#30DayMapChallenge 2024 Day 16: Choropleth",
      sources   = c("U.S. Census Bureau", "NCES"))
  ) +
  theme_rcds_map(canvas = "dark")

rcds_export(main, "day16.png",          # 5. export with a preset
            preset = "poster_land", canvas = "dark")
```

That's the whole loop. The rest of this vignette unpacks each piece.

## 2. Fonts: one voice per map

The archive swapped fonts every day, which diluted the identity. `rcds` keeps the
expressiveness but assigns fonts to **roles** (`display` / `body` / `caption`) and
ships five sanctioned **voices**. Pick one per map; never mix.

```r
rcds_fonts("default")    # Oswald / Roboto Condensed / Roboto  (everyday)
rcds_fonts("editorial")  # Anton / Roboto Condensed / Roboto   (bold posters)
rcds_fonts("vintage")    # Bebas Neue / Roboto Condensed / Roboto
rcds_fonts("fantasy")    # Cinzel / Cormorant SC / Roboto
rcds_fonts("techno")     # Orbitron / Smooch Sans / Roboto      (sports, data-art)
```

Refer to a font by role, never by Google name — that indirection is what lets you
restyle a whole map by changing the voice:

```{r font, eval = FALSE}
ggplot2::element_text(family = rcds_font("display"))
```

Sizes come from one ratio-based scale, so you never hand-tune `size = 54` again:

```{r typescale}
round(rcds_type_scale(base = 14), 1)
```

## 3. Colour: accessible families, not rainbows

The biggest upgrade over the archive. Pick a **family** by data type:

```{r palettes}
rcds_palettes()
```

```{r pal-examples}
rcds_pal("seq_blue", 5)        # sequential: low -> high
rcds_pal("div_balance", 7)     # diverging: meaningful midpoint
rcds_pal("qual_brand", 4)      # qualitative: unordered categories
```

Qualitative families refuse to over-extend — categories are capped so you don't
end up with eight indistinguishable hues:

```{r pal-cap, error = TRUE}
rcds_pal("qual_brand", 99)
```

Drop them into ggplot with the scale constructors:

```r
+ scale_fill_rcds_c("seq_amber")          # continuous
+ scale_fill_rcds_d("qual_soft")          # discrete
+ scale_color_rcds_c("div_temp", reverse = TRUE)
```

Preview every family on the dark canvas with `rcds_show_palettes()`.

## 4. Theme, signature, and composition

`theme_rcds_map()` replaces the copy-pasted theme blocks. Canvases are named:
`"dark"`, `"deep"`, `"slate"`, `"graphite"`, `"light"`, `"vintage"`.

The signature is the brand wordmark — always built the same way:

```{r signature}
cat(rcds_signature(
  challenge = "#30DayMapChallenge 2024 Day 1: Points",
  sources   = c("NCES", "U.S. Census Bureau"),
  handle    = "@rorich"))
```

Assemble a main map, a locator inset, and a credits strip with one call:

```{r compose, eval = FALSE}
loc <- rcds_locator(context = us_states, highlight = georgia)
cr  <- rcds_credits("#30DayMapChallenge 2024 Day 16: Choropleth",
                    sources = "U.S. Census Bureau", align = "right")

fig <- rcds_compose(main, locator = loc, credits = cr, layout = "poster")
rcds_export(fig, "day16.png", preset = "poster_land", canvas = "dark")
```

## 5. Score it

Rate a finished map on the ten weighted criteria (0–1 each) to get a 100-point
score, a letter grade, and a breakdown — old and new maps on the same yardstick:

```{r score}
res <- rcds_score(
  layout = 0.85, hierarchy = 0.8, typography = 0.9, colour = 0.7,
  accessibility = 0.6, legends = 0.8, labelling = 0.6, balance = 0.85,
  storytelling = 0.7, technical = 0.9, map = "draft")
res$score
res$grade
```

`rcds_score_template()` lists every criterion, weight, and what it asks for.

## 6. Lint before you export

`rcds_lint()` enforces the design system's anti-patterns mechanically, so the old
habits don't creep back:

```{r lint}
bad <- '
ggplot(d) + geom_sf(aes(fill = v), size = 0.2) +
  scale_fill_distiller(palette = "Spectral")
ggsave("m.png", width = 10, height = 8)
'
rcds_lint(bad)[, c("rule", "severity", "line", "fix")]
```

Scan a whole folder with `rcds_lint_dir("archive")`.

## Where to go next

- `docs/RCDS.md` — the full standard (principles, brand guide, pattern library).
- `inst/templates/` — runnable skeletons for choropleth, points, proportional
  symbols, bivariate, and small-multiple maps.
- `docs/innovation-log.md` — techniques to grow into (hillshade, uncertainty,
  animation, CVD proofing).
