---
title: "Untitled"
---

### Updated Elevation and School Locations Map Project

Here is the updated version of your project using ggplot2 for visualization. The previous steps for obtaining the elevation and school layers are now assumed complete, focusing only on rendering the layers using ggplot2.

#### Required Libraries

```{r}
library(sf)
library(ggplot2)
library(raster)
library(httr)
library(rvest)
library(dplyr)
library(reticulate)
library(elevatr)
```

#### Load Elevation Data and School Data

```{r}
# Load Elevation Data
georgia_bbox <- data.frame(x = c(-85.6052, -80.8397), y = c(30.3567, 35.0008))  # Rough bounding box for Georgia

# Assume elevation_data is already retrieved using get_elev_raster or from a local DEM file
elevation_data <- get_elev_raster(locations = georgia_bbox, prj = "EPSG:4326", z = 5)  # z specifies zoom level; adjust accordingly

# Convert raster to data frame for ggplot
elevation_df <- as.data.frame(elevation_data, xy = TRUE)

# Remove NA values
elevation_df <- na.omit(elevation_df)

# Rename columns
elevation_df <- elevation_df %>% rename(elevation = file1b8453a95ce1)

# Load School Data
source("nces_data_io.R")

lvl <- c(`Elementary` = "Elementary", `Middle` = "Middle", `High` = "High")

ga_schools <- ga_schools |> mutate(LEVEL = recode_factor(LEVEL, !!!lvl))

ga_schools <- ga_schools |> arrange(LEVEL)

# Convert school data to sf object
school_data <- ga_schools %>% filter(!is.na(LAT) & !is.na(LON))  # Adjust column names as needed

schools_sf <- st_as_sf(school_data, coords = c("LON", "LAT"), crs = 4326)
```

#### Plotting the Elevation and School Data with ggplot2

```{r}
# Plot elevation using ggplot
p <- ggplot() +
  geom_raster(data = elevation_df, aes(x = x, y = y, fill = elevation)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Elevation Map of Georgia",
       x = "Longitude", y = "Latitude", fill = "Elevation (m)")

# Overlay school points
p +
  geom_point(data = schools_sf, aes(x = st_coordinates(schools_sf)[, 1], y = st_coordinates(schools_sf)[, 2], color = LEVEL), size = 2, alpha = 0.8) +
  scale_color_manual(values = .tm10equal) +
  labs(color = "School Level") +
  theme(legend.position = "bottom")
```
